Computational methods in landscape ecology

### 2024-09-19 Manuscript.qmd
Minor cosmetic changes

keywords: spatial data, spatial patterns, open-source, scientific software

@@

@@

Landscapes can be defined as mosaics of different land covers, habitats, ecosystems, or land-use systems. 
The link between spatial heterogeneity and ecological processes is the core concept in the research field of landscape ecology. 
Due to its cross-disciplinary nature, the increasing availability of data, and the complexity of landscape systems, advanced computational methods are essential.

@@

This review provides an overview of recent developments in computational methods that have advanced landscape ecology research. 
We focus on key topics such as spatial patterns, connectivity, simulations and modeling, landscape genetics, and spatial planning.

@@

We are thankful to the anonymous reviewers and the Editors who provided valuable comments to improve the manuscript.

>>>>

@@

\noindent Keywords: (1) spatial data, (2) spatial patterns, (3) pattern-process link (4) open-source, (5) scientific software

@@

Landscapes can be defined as mosaics of different land covers, habitats, ecosystems, or land-use systems.
The link between spatial heterogeneity and ecological processes is the core concept in the research field of landscape ecology.
Nowadays, advanced computational methods are essential to the field due to its cross-disciplinary nature, the increasing availability of data, and the complexity of landscape systems.

@@

This review provides an overview of recent developments in computational methods that have advanced the research field of landscape ecology.
We focus on key topics such as spatial patterns, connectivity, simulations and modelling, landscape genetics, and spatial planning.

@@

We are thankful to the anonymous reviewers and the editors who provided valuable comments to improve the manuscript.

### 2024-09-19 Sections/01-introduction.qmd
Remove trailing whitespaces
Minor cosmetic changes

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modeling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2002KeyIssues; @Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

Here, we aim to introduce the latest developments of computational methods in landscape ecology.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them. 
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them.
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
![Overview of landscape ecology topics for which computational developments are discussed.](Fig-1.png)

### 2024-09-19 Sections/02-spatial-patterns.qmd
Minor cosmetic changes

Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified regarding their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
This is in line with various reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency, which are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->

@@

The Landscape Mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes.
The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class. 
This allows the assessment of content, context, and interface zones of LULC classes in a scale-dependent manner. 
Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].

@@

<!-- JN: ask T. Stepinski for review of this section later -->
<!-- MH: Maybe we need to shorten this a bit -->

@@

The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape omitting the spatial configuration.
Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances [@Claramunt2012SpatiotemporalForm].
However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
@Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix, which could be used to calculate various entropy measures.
This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).

Based on notions from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
@Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on raster's surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [e.g., @Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial]

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

Entropy metrics are available in various software such as R's *landscapemetrics*, *bespatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], *LandComp* [@Konrad2023AdaptingMultiscale], or Python's *PyLandStats*.

@@

Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm] and were adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].

Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].

@@

Many landscape ecology studies are based on the raster data model.
However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
For example, many LULC maps are provided as raster as they often related to underlying remote sensing products.
Yet, nowadays, vector data is increasingly available, and most geospatial software supports it.

Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022VecLIFramework].
At the same, the vector data model has been widely used in other fields, such as geography or urban studies.
For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].

Earlier approaches of quantifying shape complexity  included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness]

Vector-based metrics also have their limitations.
The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
Another technical issue is the need for the data to be topologically correct (e.g., geometries cannot overlap), which is often problematic especially for data from different sources.
Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.

The *vectormetrics* R package [@Matuszek2024VectormetricsLandscape] is specifically designed for vector metrics.
However, all the general software frameworks mentioned earlier, which support vector data handling, can also be used to calculate basic metrics.

@@

Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping.
These operations are based on spatial signatures which are multi-numerical representations of landscape pattern and dissimilarity measures which are functions that quantify differences between the signatures.
Spatial patterns can be compared between different areas, or for the same area at two different moments in time.

@@

R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.

>>>>

Traditionally and still the most prominent approaches to quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified. 
This includes shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address some limitations, such as correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify core metrics that capture main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
This is in line with several reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency.
These are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->
<!-- MH: I moved sections around quite a bit unfortunately -->

@@

The Landscape Mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes.
The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class.
This allows the assessment of content, context, and interface zones of LULC data in a scale-dependent manner.
Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].

@@

<!-- JN: ask T. Stepinski for review of this section later -->

@@

The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape omitting the spatial configuration.
Shannon's entropy can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and interclass distances [@Claramunt2012SpatiotemporalForm].
However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
@Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix, which could be used to calculate various entropy measures.
This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).

Based on notions from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
@Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [e.g., @Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial].

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the landscape complexity.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases, when the dissimilarities between LULC classes are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

Entropy metrics are available in various software such as the R packages *landscapemetrics*, *bespatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], *LandComp* [@Konrad2023AdaptingMultiscale], or the Python package *PyLandStats*.

@@

Surface metrics are based on the gradient surface model using gridded raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm]. 
These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
The gradient surface model can increase the resemblance of the data and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].

Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and related textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values as implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].

@@

Many landscape ecology studies are based on the raster data model.
However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products
Yet, nowadays, vector data is increasingly available and has been widely used in other fields, such as geography or urban studies.

Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022VecLIFramework].
For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
Earlier approaches to quantify shape complexity included four categories of compactness measures, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].

However, also vector-based metrics have their limitations.
The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap), which is often problematic especially for data from different sources.
Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.

The *vectormetrics* R package [@Matuszek2024VectormetricsLandscape] is specifically designed for vector metrics.
However, all general software frameworks designed for spatial data handling which support vector data can theoretically be used to calculate basic vector metrics.

@@

Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping.
These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures).
Spatial patterns can be compared between different areas, or for the same area at two different moments in time.

@@

R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.
